---
title: "Create overview on the full ARCHS4 dataset and check how well we can
assign GTEx tissues"
output:
  html_document:
    fig_width: 10
---

```{r libraries}
knitr::opts_chunk$set(echo = FALSE, fig.align="center",
                      message = FALSE, warning=FALSE)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(knitr)
library(pheatmap)
library(Rtsne)

expr <- read_tsv(snakemake@input$expr) %>%
  rename(gene_name = X1)

design <- read_tsv(snakemake@input$design) %>%
  mutate(tissue = gsub("Homo sapiens |Human","",tissue))

# sanity check for sample ordering
print("Order of samples in expr and design is the same:")
all(colnames(expr %>% mutate(gene_name = NULL)) == design$sample)

```
# Inroduction

This is an exploratory analysis of the downloaded and processed ARCH4 data for
the tissue keywords "`r snakemake@wildcards$keywords`".
These data consists of `r nrow(expr)` genes measured in `r ncol(expr)-1` samples.

Let's check the expression and corresponding design tables:

```{r load_data}

print("Expression:")
kable(head(expr))

print("Design:")
kable(head(design))
```

The design table contains the following column names:
```{r design_cols}
colnames(design)
```

We now have a look at the individual columns, specifically what kind of values
we can find.
Below we show for each column of the design matrix the unique values contained
therein.

```{r values_by_cols}
lapply(colnames(design), function(x) {
  design %>% select(one_of(x)) %>% unique
})
```

Let's also check how many samples we get if we group by the individual values
for each of the columns:

```{r samples_by_groups}
# we remove id, series and description
cols <- colnames(design)[-c(1,3,5)]

lapply(cols, function(x) {
  design %>% group_by_(.dots = as.symbol(x)) %>%
    summarize(count=n()) %>%
    arrange(desc(count))
})

```

# Gene expression

Ok, now let's have a look at the actual gene expression data.
We will have a look at the histogram of all expression values and a heatmap
of the top 1% most variable genes.

```{r expression_overview}
expr_subs <- expr
expr_subs$gene_name <- NULL
expr$gene_var <- apply(expr_subs,1,var)
expr_subs <- filter(expr, gene_var > quantile(gene_var, .99))

# histogram
toplot <- expr %>% mutate(gene_var = NULL, gene_name = NULL) %>%
  melt
ggplot(toplot, aes(x=value)) + geom_histogram() + xlab("expression") +
  ggtitle("Distribution of expression values.")

# heatmap
expr_corr <- mutate(expr_subs, gene_var = NULL, gene_name = NULL) %>% t %>% cor
pheatmap(expr_corr, main = "Expression correlations of top 1% variable genes.")

```

# t-SNE
Now let's do some tSNE plots to see whether we can see any specific clusters emerging.

```{r tsne}

expr_subs <- t(mutate(expr, gene_var = NULL, gene_name = NULL))

# tsne
reduction <- Rtsne(expr_subs, max_iter = 500, theta = 0.0,
                   dims = 2, perplexity = 30)

# plotting
toplot <- data.frame(reduction$Y)
colnames(toplot) <- c("dim1", "dim2")

toplot <- cbind(toplot, tissue = design$tissue, instrument = design$instrument,
                series=design$series)

ggplot(toplot, aes(x=dim1, y=dim2, col=tissue)) +
  geom_point() +
  ggtitle("t-SNE on gene expression data labeled \nby 'tissue' meta-data information")

ggplot(toplot, aes(x=dim1, y=dim2, col=instrument)) +
  geom_point() +
  ggtitle("t-SNE on gene expression data labeled \nby 'instrument' meta-data information")

ggplot(toplot, aes(x=dim1, y=dim2, col=series)) +
  geom_point() +
  ggtitle("t-SNE on gene expression data labeled \nby 'Series' meta-data information")

```


## Session Info
```{r}
sessionInfo()
```
