---
title: "Exploration of ARCHS4 gene expression data"
output:
  html_document:
    fig_width: 10
---

```{r libraries}
# set options and load libraries
knitr::opts_chunk$set(echo = FALSE, fig.align="center", 
                      message = FALSE, warning=FALSE)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(knitr)
library(pheatmap)
library(Rtsne)

# read the expression data (raw and normalized
expr <- read_tsv(snakemake@input$expr) %>%
  rename(gene_name = X1)
raw <- read_tsv(snakemake@input$raw) %>%
  rename(gene_name = X1)

# load design and do some cleaning
# NOTE: we only download the H. sapiens file
design <- read_tsv(snakemake@input$design) %>%
  mutate(tissue = trimws(tolower(gsub("Homo sapiens |Human","",tissue))))

# sanity check for sample ordering
print("Order of samples in expr and design is the same:")
all(colnames(expr %>% mutate(gene_name = NULL)) == design$sample)

```
# Inroduction

This is an exploratory analysis of the downloaded and processed ARCH4 data for
the tissue keywords "`r snakemake@wildcards$keywords`".
These data consists of `r nrow(expr)` genes measured in `r ncol(expr)-1` samples.

Let's check the expression and corresponding design tables:

```{r load_data}

print("Expression:")
kable(expr[1:5,1:10])

print("Design:")
kable(head(design))
```

The design table contains the following column names:
```{r design_cols}
colnames(design)

# for now we quit here if we didnt get any data
if(nrow(expr)<2) quit("no",0)
```

We now have a look at the individual columns, specifically what kind of values
we can find.
Below we show for each column of the design matrix the unique values contained
therein. This is a bit messy, but for now we accept this.

```{r values_by_cols}
lapply(colnames(design), function(x) {
  design %>% select(one_of(x)) %>% unique
})
```

Let's also check how many samples we get if we group by the individual values
for each of the columns:

```{r samples_by_groups}
# we remove id, series and description
cols <- colnames(design)[-c(1,3,5)]

counts <- lapply(cols, function(x) {
  design %>% group_by_(.dots = as.symbol(x)) %>% 
    summarize(count=n()) %>% 
    arrange(desc(count))
})
names(counts) <- cols

# get the tissue counts and match
# with original design
tissue_counts <- counts$tissue
matched_counts <- tissue_counts[match(design$tissue, tissue_counts$tissue),]
design <- rename(design, tissue_old = tissue) %>%
  mutate(tissue = paste(matched_counts$tissue, " (", matched_counts$count, ")", sep=""))

# report the tissue counts
kable(counts$tissue)
```

# Raw and batch normalized gene expression

Ok, now let's have a look at the actual gene expression data.
We will have a look at the histogram of all expression values and a heatmap
of the top 1% most variable genes.

```{r expression_overview}
expr_subs <- expr
expr_subs$gene_name <- NULL
expr$gene_var <- apply(expr_subs,1,var)
expr_subs <- filter(expr, gene_var > quantile(gene_var, .99))

raw_subs <- raw
raw_subs$gene_name <- NULL
raw$gene_var <- apply(raw_subs,1,var)
raw_subs <- filter(raw, gene_var > quantile(gene_var, .99))

# histogram
toplot <- expr %>% mutate(gene_var = NULL, gene_name = NULL) %>%
  melt
ggplot(toplot, aes(x=value)) + geom_histogram() + xlab("expression") +
  ggtitle("Distribution of expression values.")

# heatmap raw
raw_corr <- mutate(raw_subs, gene_var = NULL, gene_name = NULL) %>% t %>% cor
pheatmap(raw_corr, main = "Correlation map of top 1% variable genes (raw).")

# heatmap norm
expr_corr <- mutate(expr_subs, gene_var = NULL, gene_name = NULL) %>% t %>% cor
pheatmap(expr_corr, main = "Correlation map of top 1% variable genes (norm).")

```

# t-SNE
Now let's do some tSNE plots to see whether we can see any specific clusters emerging.
Specifically, we will look at two types of t-SNE plots: One for the raw expression data as 
was extracted from ARCHS4 and one using the batch normalized data.

```{r tsne}

expr_subs <- t(mutate(expr, gene_var = NULL, gene_name = NULL))
raw_subs <- t(mutate(raw, gene_var = NULL, gene_name = NULL))

# tsne
reduction <- Rtsne(expr_subs, check_duplicates=FALSE, max_iter = 1000, theta = 0.0, 
                   dims = 2, perplexity = 30)
reduction_raw <- Rtsne(raw_subs, check_duplicates=FALSE, max_iter = 1000, theta = 0.0, 
                   dims = 2, perplexity = 30)

# plotting
toplot <- data.frame(rbind(reduction$Y, reduction_raw$Y))
colnames(toplot) <- c("dim1", "dim2")

toplot <- mutate(toplot, type = c(rep("normalized", nrow(reduction$Y)),
                                  rep("raw", nrow(reduction_raw$Y))))

toplot <- cbind(toplot, tissue = design$tissue, instrument = design$instrument,
                series=design$series)

gp1 <-ggplot(toplot, aes(x=dim1, y=dim2, col=tissue)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'tissue' meta-data information")

gp2 <- ggplot(toplot, aes(x=dim1, y=dim2, col=instrument)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'instrument' meta-data information")

gp3 <- ggplot(toplot, aes(x=dim1, y=dim2, col=series)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'Series' meta-data information")

gp1
gp2
gp3

# we also save these plots separtely to a pdf
pdf(snakemake@params$pdf_out, width=15, height=10)
gp1
gp2
gp3
dev.off()
```


## Session Info
```{r}
sessionInfo()
```
